Cybersecurity impacts at the beginning\\ of Covid-19 pandemic in Italy

### 2021-03-06 article.tex
Working on draft

% Institutes for affiliations are also joined by \and,
\institute{
  Associazione Italiana Professionisti della Sicurezza Informatica (AIPSI), Milano, Italy
\and
  Italian Information System Security Association International (ISSA)
\and
	Digital Attacks Observatory Team (OAD), Italy
\and
   University of Verona, Verona, Italy\\
   \email{\{fausto.spoto,luca.olivieri\}@univr.it}\\
 }

@@

The advent of the Covid-19 pandemic and the consequent lock-downs have led companies to face needs such as smart working, remotely work and digital business,
accelerating their digitalization efforts. As reported in \cite{Goasduff20} by Gartner, a world's leading research and advisory company, before the pandemic crisis
most organizations moved their digital strategies forward at a steady pace, but during the Covid-19 pandemic there has been a significant increase in the 
development of digital products and services to maintain and accelerate customer engagement. However, the digitalization race has led to many cyber security 
issues and the intensification of malicious cyber attacks around the world~\cite{HKICS20,PA21}. In this scenario, Italy is an interesting case of study because 
it is composed mostly from Small Medium Enterprises (SMEs) and therefore can highlight specific aspects instead of the various global views in which large companies 
have a greater influence.

This paper presents a window of italian panorama to understand how the cyber-security is faced up during COVID-19 pandemic. 
Section~\ref{sec:DigitalAttacksObservatory} presents \textit{Digital Attacks Observatory}, a survey containing relevant information about cyber security and 
italian organizations which will be used in subsequent sections as a reference to better understand Covid-19 impacts. A detailed discussions of results are in 
Sections~\ref{sec:DataDiscussion,sec:TypeAttacks}. Finally, Section~\ref{sec:Conclusion} concludes by underlying the importance of cybersecurity culture focusing 
on Italian reality and SMEs.

@@

The OAD questionaries refer for about half of the question to the cyber attacks typologies and they main characteristics, if they have been relived. 
The other half of the questions refers to the technical and organizational security measures implemented on the IT systems of the respondents. 
OAD therefore allows to contextualize the cyber attacks with the cyber measures in production in each IT system considered. 
In fact, OAD provided a macro qualitative evaluation of the implemented measures (within the specific context of the company), in order to improve the motivation to 
complete the questionaries to all the potential anonymous respondents. 
In particular the security measures considered in the 2020 questionnaire are subdivided in technical, organizational, management and governance ones as follow:
\begin{enumerate}
\small
	\item \emph{Technical measures}
	\begin{itemize}
		\item Overall architecture of digital security measures, integrated with the entire IT system architecture
		\item Physical countermeasures
		\item Identification, Authentication, and Authorization (IAA) measures
		\item Cyber security measures at local network and geographic level
		\item Measures for logical protection of each single Information and Communications Technology (ICT) system
		\item Application and software protection
		\item Data protection
	\end{itemize}
	\item \emph{Organizational measures}
	\begin{itemize}
		\item Organizational structure, roles, skills, certifications
		\item Organizational policies and procedures
		\item Contracts, agreements and digital security clauses with third parties
		\item Awareness and culture of digital security at all levels
		\item Auditing
	\end{itemize}
	\item \emph{Management and governance measures}
	\begin{itemize}
		\item Digital security control, monitoring and management systems
		\item Disaster Recovery (DR) plan
	\end{itemize}
\end{enumerate}

@@

OAD 2020 survey\footnote{The report is written in Italian, only the Executive Summary is in English: 186 A4 pages, 148 images and graphics, 11 Chapters
(147 A4 pages) and 9 Attachments (39 A4 pages)}~\cite{oad20} took place during the Covid-19 pandemic, and it covered the entire 2019 and the first quarter 
of 2020, when in Italy the pandemic exploded. The data comparison of 2019 with these of the first quarter 2020, provided by the same pool of respondents, has a statistical relevance and provides a 
clear indication of the Covid-19 pandemic on the intentional cyber attacks in Italy. This pandemic has been the trigger for a wide range of cyber attacks, mainly 
caused by the sudden - and in part unprepared - passage of many to agile working from home and to a strong use of every type of IT service on Internet, mainly
derived by the mobility lockdowns imposed by the Italian Authorities. The emerged respondents’ pool covers all the product sectors, including Public Administrations,
even if the majority of the respondents’ companies belong to the ICT sector (30.8\%). The 2020 pool  is well balanced for the size of the organizations, in terms
of number of employees, between those below 250 and the largest ones. It must be taken into account that in Italy 99.91\% \cite{istat20} %TODO: manca fonte/citazione
of enterprises are SMEs, under 250 employees, and 95\% are under ten employees. The OAD 2020 survey is therefore able to consider small and 
tiny organizations, which are thee vast majority in Italy, and which are not normally considered in other cyber security surveys: 57.5\% of the OAD 2020 respondents
belong to structures with less than 250 employees, and of these 22.1\% under 10.

@@

\begin{figure}
	\centering
		\includegraphics[width=1\textwidth]{pictures/fig3.png}
		\caption{Comparison between cyber attacks relived in 2019 and first quarter of 2020 for the organization dimension~\cite{oad20}. The black dotted line in the figure shows 
the “not” relieved attacks, and points out how it decrease from the tiny to the big organizations, in terms of numbers of employees. }
		\label{fig:3}
\end{figure}

@@

\begin{table}[h]
  \begin{center}
  \resizebox{\textwidth}{!}{
	\begin{tabular}{|r|r|r|r|r|r|}									
		\hline									
		Critical structure protection	&	1 Jan - 30 Apr 2020	&	1 Jan - 31 Dec 2019	&	1 Jan - 31 Dec 2018 &	1 Jan - 31 Dec 2017 &	1 Jan - 31 Dec 2016\\
		\hline									
		Relevant attacks	&	282	&	1.181	&	459	&	1.032 & 844	\\
		Issued alerts	&	24.824	&	82.484	&	80.777	&	31.524 & 6.721	\\
		Initiated investigations 	&	34	&	155	&	74	&	72 & 70	\\
		People arrested	&	0	&	3	&	1	&	3	& 3\\
		People reported	&	0	&	117	&	14	&	1.316 &	1.226	\\
		Perquisitions	&	n.d.	&	n.d.	&	n.d.	&	73	& 58 \\
		\hline	
		Initiated investigations on detected attacks	&	12,05\%	&	13,12\%	&	16,12\%	&	6,98\%	& 8,29\% \\
		\hline	
		People arrested out of	reported &	0\%	&	2,5\%	&	7,14\%	&	0,23\%	&	0,24\%\\						
		\hline	
	\end{tabular}
	}
    \end{center}

@@

\begin{table}[h]
  \begin{center}
  \resizebox{\textwidth}{!}{
	\begin{tabular}{|r|r|r|r|r|r|}									
		\hline									
		Financial Cyber Crime	&	1 Jan - 30 Apr 2020	&	1 Jan - 31 Dec 2019	&	1 Jan - 31 Dec 2018 &	1 Jan - 31 Dec 2017 &	1 Jan - 31 Dec 2016\\
		\hline									
		Blocked fraudulent transactions	&	\textbf{\euro 20.200.000,00}	&	\textbf{\euro 21.333.990,00}	&	\euro 38.400.000,00	& \euro 20.839.576,00 & \euro 16.050.812,50 	\\
		Recovered amounts 	&	\euro 8.700.000,00	&	\euro 18.000.000,00	&	\euro 9.000.000,00	&	\euro 862.000,00 & n.d.	\\
		People arrested	&	n.d.	&	n.d.	&	n.d.	&	25	& 25\\
		People reported	&	n.d.	&	n.d.	&	n.d.	&	2.851 & 3.772	\\
		\hline	
		Recovered amounts out of fraud	&	12,05\%	&	13,12\%	&	16,12\%	&	6,98\%	& 8,29\% \\
		\hline	
	\end{tabular}
	}
    \end{center}
	\caption{Data of critical structure protection collected by Italian Postal and Communications Police~\cite{oad20}}									
	\label{tab:table2}									
\end{table}

@@

The type of attack (what is attacked) most widespread in both the considered periods is the one aimed at access control systems (IAA, Identification Authentication Authorization), 
with 34\% in 2019 and 28.2\% in the 1st quarter of 2020, as shown in Fig.~\ref{fig:5}. As diffusion among the respondents, the other 14 types of attacks follow, decreasing by a few percentage
points among them, whose characteristics and impacts are detailed in the related paragraphs of the OAD 2020 report. Referring to the attack techniques (how to attack), all the seven
attacks families considered in the 2020 questionnaire are widely used in the various types of attacks, some times even simultaneously., as shown in Fig.~\ref{fig:6}. The most widespread one,
as an average on the various attacks detected by the respondents’ pool, is the use of toolkits (rootkits, meta-exploits, etc.) for the identification and exploitation of vulnerabilities
on the target system of the attack, with a 38.8\%, followed by the well-known malicious and unauthorized collection of information (social engineering, phishing, etc.) with 34.6\% 
and the use of malicious code and scripts with 34.3%. The other considered technolog decrease with a few percentage points of difference between them. The impacts of the most 
critical attacks are analyzed for each attack type, as well as their possible motivations and the recovery times required by the most critical ones.

In the 2020 OAD survey, all these attacks’ characteristics vary for each type of attack, and the emerged results are described in the specific paragraphs dedicated to each attack 
type. At overall it emerges that:
\begin{itemize}
\item the impacts declared by the respondents are balanced between those irrelevant and / or easily resolvable with recovery in short time and at limited costs, and those very 
critical, which require expensive and long recovery actions, and that, in some cases, can cause business and customer loses; these two very different impact cases depend mainly 
on the security measures in place;

@@

As pointed out in Fig.~\ref{fig:1}, the Informatics Systems considered in the 2020 survey and their cybersecurity  are in the medium-high range and the emerged information show a relevant 
improvement in both technical and organizational measures, in comparison to the previous surveys. Few of the considered Italian Informatics Systems are based on Data Centers in 
Italy, most of the respondents’ companies have medium-small Informatic Systems partly on premise and partly outsourced, with an increasing use of cloud services.
The high number of attacks detected in 2018, and the privacy obligations, related to General Data Protection Regulation (GDPR), have certainly contributed to strengthening cyber security measures, and a further improvement 
comes from the growing use of cloud services, where usually there are high standard security measures. Organizational measures for cybersecurity, historically lacking and neglected
in Italy, have improved in terms of defining roles and separation of duties, of organizational policies and procedures, and of incidents management. These measures often are lacking
in small organizations, and in general the cybersecurity awareness and competence are low, and mainly at the top level of the public and private organizations. In Italy, there is 
still a long way to go in terms of cybersecurity continual training and awareness. A formal and bureaucratic approach for the organizational procedures is still prevalent, which 
once defined, often do not find a possible concrete applicability, periodic updates and operational tests: an emblematic example is represented by the Disaster Recovery plans for 
which, several times, the required alternative ICT resources to be used are not forecasted and allocated and there also not provided periodic exercises and simulations of the 
possible disasters. Cybersecurity management tools still have limited diffusion among the 2020 OAD respondents, in particular the most advanced ones based on artificial intelligence
techniques. The use of IoT, of industrial automation and of robotics systems, and also of systems based on blockchain technologies find a low number of respondents involved, which 
also derives from the limited percentage of their product sectors that should be the more interested in these systems: the manufacturing sectors, the logistics, the research and 
development centers and labs, the Local Public Administrations for territorial control, and so on. For these topics, the data that emerged from the survey are considered only as 
specific cases that contribute to the general values on the cyberattacks, but which cannot be still considered of reference at the Italian level.

>>>>

% Institutes for affiliations are also joined by \and,
\institute{
  Associazione Italiana Professionisti della Sicurezza Informatica (AIPSI), Milano, Italy
\and
  Italian Information System Security Association International (ISSA)
\and
	Digital Attacks Observatory (OAD) Team, Italy
\and
   University of Verona, Verona, Italy\\
   \email{\{fausto.spoto,luca.olivieri\}@univr.it}\\
 }

@@

The advent of Covid-19 pandemic and the consequent lock-downs have led companies to face needs such as smart working, remotely work and digital business,
accelerating their digitalization efforts. As reported in \cite{Goasduff20} by Gartner, a world's leading research and advisory company, before the pandemic crisis
most organizations moved their digital strategies forward at a steady pace, but during the Covid-19 pandemic there has been a significant increase in the 
development of digital products and services to maintain and accelerate customer engagement. However, the digitalization race has led to many cyber security 
issues and the intensification of malicious cyber attacks around the world~\cite{HKICS20,PA21}. In this scenario, Italy is an interesting case of study because 
it is composed mostly from Small Medium Enterprises (SMEs) and therefore can highlight specific aspects instead of the various global views in which large companies 
have a greater influence. This paper presents a window of italian panorama to understand how the cyber-security is faced up during COVID-19 pandemic. 
Section~\ref{sec:DigitalAttacksObservatory} presents \textit{Digital Attacks Observatory}, a survey containing relevant information about cyber security and 
italian organizations which will be used in subsequent sections as a reference to better understand Covid-19 impacts. A detailed discussions of collected data are in 
Sections~\ref{sec:DataDiscussion}, \ref{sec:TypeAttacks}. Finally, Section~\ref{sec:Conclusion} concludes by underlying the importance of cybersecurity culture focusing 
on Italian reality and SMEs.

@@

The OAD questionaries refer for about half of the question to cyber attacks typologies and they main characteristics, if they have been relived. 
The other half of the questions refers to the technical and organizational security measures implemented on the IT systems of the respondents. 
OAD therefore allows to contextualize the cyber attacks with the cyber measures in production in each IT system considered. 
In fact, OAD provided a macro qualitative evaluation of the implemented measures (within the specific context of the company), in order to improve the motivation to 
complete the questionaries to all the potential anonymous respondents. 
In particular the security measures considered in the 2020 questionnaire are subdivided in technical, organizational, management and governance ones as follow:
\begin{enumerate}
\small
	\item \emph{Technical measures}
	\begin{itemize}
		\item Overall architecture of digital security measures, integrated with the entire IT system architecture
		\item Physical countermeasures
		\item Identification, Authentication, and Authorization (IAA) measures
		\item Cyber security measures at local network and geographic level
		\item Measures for logical protection of each single Information and Communications Technology (ICT) system
		\item Application and software protection
		\item Data protection
	\end{itemize}
	\item \emph{Organizational measures}
	\begin{itemize}
		\item Organizational structure, roles, skills, certifications
		\item Organizational policies and procedures
		\item Contracts, agreements and digital security clauses with third parties
		\item Awareness and culture of digital security at all levels
		\item Auditing
	\end{itemize}
	\item \emph{Management and governance measures}
	\begin{itemize}
		\item Digital security control, monitoring and management systems
		\item Disaster Recovery (DR) plan
	\end{itemize}
\end{enumerate}

@@

OAD 2020 survey\footnote{The report is written in Italian, only the Executive Summary is in English: 186 A4 pages, 148 images and graphics, 11 Chapters
(147 A4 pages) and 9 Attachments (39 A4 pages)}~\cite{oad20} took place during the Covid-19 pandemic, and it covered the entire 2019 and the first quarter 
of 2020, when in Italy the pandemic exploded. The data comparison of 2019 and the first quarter 2020, provided by the same respondents pool, has a statistical relevance and provides a 
clear indication of the Covid-19 pandemic on the intentional cyber attacks in Italy. This pandemic has been the trigger for a wide range of cyber attacks, mainly 
caused by the sudden - and in part unprepared - passage of many to agile working from home and to a strong use of every type of IT service on Internet, mainly
derived by the mobility lockdowns imposed by the Italian Authorities. The emerged respondents’ pool covers all the product sectors, including Public Administrations,
even if the majority of the respondents’ companies belong to the ICT sector (30.8\%). The 2020 pool is well balanced for the size of the organizations, in terms
of employees, between those below 250 and the largest ones. It must be taken into account that in Italy 99.91\% \cite{istat20} %TODO: manca fonte/citazione
of enterprises are SMEs, under 250 employees, and 95\% are under ten employees. The OAD 2020 survey is therefore able to consider small and 
tiny organizations, which are thee vast majority in Italy, and which are not normally considered in other cyber security surveys: 57.5\% of the OAD 2020 respondents
belong to structures with less than 250 employees, and of these 22.1\% under 10.

@@

\begin{figure}
	\centering
		\includegraphics[width=1\textwidth]{pictures/fig3.png}
		\caption{Comparison between cyber attacks relived in 2019 and first quarter of 2020 for the organization dimension~\cite{oad20}. The black dotted line in the figure shows 
the “not” relieved attacks, and points out how it decrease from the tiny to the big organizations, in terms of employees. }
		\label{fig:3}
\end{figure}

@@

\begin{table}[h]
  \begin{center}
  \resizebox{1.15\textwidth}{!}{
	\begin{tabular}{|r|r|r|r|r|r|}									
		\hline									
		Critical structure protection	&	1 Jan - 30 Apr 2020	&	1 Jan - 31 Dec 2019	&	1 Jan - 31 Dec 2018 &	1 Jan - 31 Dec 2017 &	1 Jan - 31 Dec 2016\\
		\hline									
		Relevant attacks	&	282	&	1.181	&	459	&	1.032 & 844	\\
		Issued alerts	&	24.824	&	82.484	&	80.777	&	31.524 & 6.721	\\
		Initiated investigations 	&	34	&	155	&	74	&	72 & 70	\\
		People arrested	&	0	&	3	&	1	&	3	& 3\\
		People reported	&	0	&	117	&	14	&	1.316 &	1.226	\\
		Perquisitions	&	n.d.	&	n.d.	&	n.d.	&	73	& 58 \\
		\hline	
		Initiated investigations on detected attacks	&	12,05\%	&	13,12\%	&	16,12\%	&	6,98\%	& 8,29\% \\
		\hline	
		People arrested out of	reported &	0\%	&	2,5\%	&	7,14\%	&	0,23\%	&	0,24\%\\						
		\hline	
	\end{tabular}
	}
    \end{center}

@@

\begin{table}[h]
  \begin{center}
  \resizebox{1.15\textwidth}{!}{
	\begin{tabular}{|r|r|r|r|r|r|}									
		\hline									
		Financial Cyber Crime	&	1 Jan - 30 Apr 2020	&	1 Jan - 31 Dec 2019	&	1 Jan - 31 Dec 2018 &	1 Jan - 31 Dec 2017 &	1 Jan - 31 Dec 2016\\
		\hline									
		Blocked fraudulent transactions	&	\textbf{\euro 20.200.000,00}	&	\textbf{\euro 21.333.990,00}	&	\euro 38.400.000,00	& \euro 20.839.576,00 & \euro 16.050.812,50 	\\
		Recovered amounts 	&	\euro 8.700.000,00	&	\euro 18.000.000,00	&	\euro 9.000.000,00	&	\euro 862.000,00 & n.d.	\\
		People arrested	&	n.d.	&	n.d.	&	n.d.	&	25	& 25\\
		People reported	&	n.d.	&	n.d.	&	n.d.	&	2.851 & 3.772	\\
		\hline	
		Recovered amounts out of fraud	&	12,05\%	&	13,12\%	&	16,12\%	&	6,98\%	& 8,29\% \\
		\hline	
	\end{tabular}
	}
    \end{center}
	\caption{Data of critical structure protection collected by Italian Postal and Communications Police~\cite{oad20}}									
	\label{tab:table2}									
\end{table}

@@

In the pandemic period between 2019 and 1st quarter of 2020, the attacks on IAA that aimed at access control systems are most widespread with 34\% in 2019 and 28.2\% (Fig.~\ref{fig:5}).
As diffusion among the respondents, the other 14 types of attacks follow, decreasing by a few percentage points among them, whose characteristics and impacts are detailed in the related 
paragraphs of the OAD 2020 report. Regarding attack vectors and techniques, all the seven attacks families considered in the 2020 questionnaire are widely used in the various 
types of attacks, some times even simultaneously., as shown in Fig.~\ref{fig:6}. The most widespread one,
as an average on the various attacks detected by the respondents’ pool, is the use of toolkits (rootkits, meta-exploits, etc.) for the identification and exploitation of vulnerabilities
on the target system of the attack, with a 38.8\%, followed by the well-known malicious and unauthorized collection of information (social engineering, phishing, etc.) with 34.6\% 
and the use of malicious code and scripts with 34.3\%. The other considered technolog decrease with a few percentage points of difference between them. The impacts of the most 
critical attacks are analyzed for each attack type, as well as their possible motivations and the recovery times required by the most critical ones.
In the 2020 OAD survey, all these attacks’ characteristics vary for each type of attack, and the emerged results are described in the specific paragraphs dedicated to each attack 
type. At overall it emerges that:
\begin{itemize}
\item the impacts declared by the respondents are balanced between those irrelevant and/or easily resolvable with recovery in short time and at limited costs, and those very 
critical, which require expensive and long recovery actions, and that, in some cases, can cause business and customer loses; these two very different impact cases depend mainly 
on the security measures in place;

@@

As pointed out in Fig.~\ref{fig:1}, the Informatics Systems considered in the 2020 survey and their cybersecurity  are in the medium-high range and the emerged information show a relevant 
improvement in both technical and organizational measures, in comparison to the previous surveys. Few of the considered Italian Informatics Systems are based on Data Centers in 
Italy, most of the respondents’ companies have medium-small Informatic Systems partly on premise and partly outsourced, with an increasing use of cloud services.
The high number of attacks detected in 2018, and the privacy obligations, related to General Data Protection Regulation (GDPR), have certainly contributed to strengthening cyber security measures, and a further improvement 
comes from the growing use of cloud services, where usually there are high standard security measures. Organizational measures for cybersecurity, historically lacking and neglected
in Italy, have improved in terms of defining roles and separation of duties, of organizational policies and procedures, and of incidents management. These measures often are lacking
in small organizations, and in general the cybersecurity awareness and competence are low, and mainly at the top level of the public and private organizations. In Italy, there is 
still a long way to go in terms of cybersecurity continual training and awareness. A formal and bureaucratic approach for the organizational procedures is still prevalent, which 
once defined, often do not find a possible concrete applicability, periodic updates and operational tests: an emblematic example is represented by the Disaster Recovery plans for 
which, several times, the required alternative ICT resources to be used are not forecasted and allocated and there also not provided periodic exercises and simulations of the 
possible disasters. Cybersecurity management tools still have limited diffusion among the 2020 OAD respondents, in particular the most advanced ones based on artificial intelligence
techniques. Although the use of IoT \cite{JK21}, industrial automation, robotics systems \cite{SGLMDXHKCZT21}, and systems based on blockchain technologies \cite{RMFF21} are expanding and can allow to combat the Covid-19 pandemic, in 
the OAD 2020, we can find only a low number of respondents involved, which also derives from the limited percentage of their product sectors that should be the more interested in these systems: the manufacturing sectors, the logistics, the research and 
development centers and labs, the Local Public Administrations for territorial control, and so on. For these topics, the data that emerged from the survey are considered only as 
specific cases that contribute to the general values on the cyberattacks, but which cannot be still considered of reference at the Italian level.
